Temporary: An Analysis of Programming Students’ Interaction Prompts with Generative AI: A Metacognitive Perspective

### 2024-09-30 CSEDU2025/Paper.tex
update paper

\title{Analysis of Programming Students' Prompts with Generative AI}

@@

\keywords{Generative Artificial Intelligence, Programming Education, Chatbots}

@@

% %What is the problem to be solved?

Recent advancements in Generative Artificial Intelligence (GenAI) have opened
up new possibilities in education. In programming courses, students can utilize
GenAI tools to improve their understanding, receiving assistance, feedback, and
detailed explanations.

The study of \cite{chan23} revealed that both undergraduate and postgraduate
students exhibit positive attitudes toward the use of GenAI in teaching and
learning, noting that it enhances the depth of their thinking and understanding.
Additionally, a systematic review conducted by \cite{Lo24} demonstrated that
students could effectively learn from ChatGPT, leading to improved understanding
and academic achievement. Besides of that, it was also observed that ChatGPT
enables students to regulate their learning pace, promoting self-regulation
\citep{Baha24} \citep{cai23}.

In the other hand, researchers have concerns about the impact of these
tools on students. The systematic review of \cite{Murillo23} indicated that
ChatGPT use might lead to an overreliance on the tool. \cite{chan23} also noted
that this reliance could result in a decrease in critical thinking, as students
might make decisions based solely on the information provided by ChatGPT.

The student's confidence is warranted, as shown by \cite{Puryear22}, who
demonstrated that GitHub Copilot can generate solutions for student assignments
with accuracy rates ranging from 68\% to 95\%. However, this raises concerns that
students might rely too heavily on GenAI tools, potentially neglecting a deeper
understanding of the underlying concepts. \cite{cai23} further identified
overdependence and reduced intellectual engagement as significant drawbacks of
using ChatGPT in learning.

Regardless of teachers' preferences or beliefs, a preliminary surveys performed
by \cite{Dickey24} revealed that at least 54.5\% of students are already using
GenAI for homework. This highlights the need to understand how students are
interacting with these tools and how they can be used to enhance learning.
Furthermore, a review by \cite{Lo24} emphasized the necessity for extended
studies and objective measures to gather more robust evidence on the use of
GenAI tools in education.

% % Are there any existing solutions?

To deal with this issue, some researchers have proposed different pedagogical
strategies \cite{Denny24} introduced the concept of \textit{Prompt
Problems}, where students solve programming exercises by formulating natural
language prompts. \cite{Prasad24} proposed a self-regulated learning framework
using GenAI to solve programming problems. \cite{Lauren23} explored the
integration of GenAI with evidence-based learning strategies in computer science
and engineering courses.

% % Which is the best?

The study of effective educational strategies in use of GenAI tools in education
context are important for determining the best training for educators. However,
as these tools gain popularity among students \cite{Dickey24} the urgency to
equip educators with best practices may lag behind their rapid adoption.
Therefore, analyzing students' interaction patterns with GenAI tools is
important for understanding how these tools are being used and how educators can
leverage them to enhance learning. In addition, these interaction patterns can
create new opportunities to redesign GenAI tools to better support pedagogical
strategies without repressing the development of abstract, critical and creative
thinking.

@@

Nevertheless, it is important to consider how GenAI tools affect students across
different demographic groups, knowledge areas, cultural backgrounds
\cite{catalan21} \cite{neo22} and students' prior experience. Thus, this works
is limited to a specific group of students and a specific GenAI tool. The
results may not be generalizable to other populations or tools.

% % What do you hope to achieve?

This study aims to analyze the interactions between students and CharlieBot,
an ChatGPT 3.5 based bot developed by UC3M that leverages Retrieval-Augmented
Generation (RAG) to enhance its contextual understanding of Java programming.
We created three research questions to guide our analysis:

RQ1: What was the distribution pattern of the classified messages?

RQ2: Was there a time gap between student submissions? If so, which category
showed the largest spacing between submissions?

RQ3: Do students’ conversations with the bot alternate between practical and
theoretical approaches, thereby promoting interleaving?

@@

@@

% Regarding self-regulation, \cite{cai23} noted that ChatGPT allowed students to
% regulate their learning pace.
\begin{table*}[htbp]
  \caption{Categories - adapted from \cite{Ghimire24}}
  \begin{center}
    \renewcommand{\arraystretch}{1.6} % Increase the spacing between rows
    \begin{tabular}{p{3cm} p{12cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Description} \\
      \hline
      Debugging Help & Prompts that seek help to identify, fix errors, or understand the provided code snippet. \\
      Conceptual Question & Prompts that are more about understanding concepts than specific code. \\
      Student Correction & Prompts where the student corrects the bot. \\
      Code Snippet & Prompts that ask for a specific part of the code, like a function or a segment. \\
      Complete Solution & Prompts that request an entire solution or a complete code snippet. \\
      Multiple Question & Prompts where the user wants to solve a multiple choice exercise (Quiz). \\
      Language Change & Prompts that request a change of idiom. \\
      Uncategorized & Prompts that do not fit into any of the above categories. \\
      \hline
    \end{tabular}
    \label{tab:categories}
  \end{center}
\end{table*}

@@

Metacognitive knowledge includes beliefs about variables that affect the
outcomes of cognitive activities. It is divided into three types: beliefs about
personal abilities, perceived difficulty in the task, and previously used
strategies. Metacognitive Experience refers to the feelings that arise
before, during, and after cognitive activity, such as frustration, confusion,
and satisfaction, among others. Metacognitive goals are the key to regulating
thought as they relate to the goals the individual seeks to achieve, directly
influencing the actions taken. For example, if a student’s goal is to complete
a task quickly, they may adopt a more passive approach to learning. Lastly,
Metacognitive Actions involve the planning, monitoring, and evaluation of
strategies used to achieve the goals. In terms of planning, students can
determine how to approach a task, such as spacing their study sessions
\citep{Ouhao18, Carvalho20}, interleaving topics \citep{Rivers21}, utilizing
retrieval practice \citep{larsen18}, and other strategies.

@@

In the categorization phase, the students' messages were manually classified
into eight distinct categories for analysis. Initially, \cite{Ghimire24}
proposed four categories: Debugging Help, Conceptual Question, Code Snippet,
and Complete Solution. However, the data collected indicated the need for
additional categories, resulting in the inclusion of four more: Multiple Choice
(Quiz), Student Corrections, Language Change, and Uncategorized. Table
\ref{tab:categories} presents these categories along with their respective
descriptions.

The analysis phase involved the use of Python scripts in the Jupyter Notebook
environment to extract relevant information from previously classified
data. The datasets and code snippets used in this process are available at: YYY.

@@

The results of the analysis are presented in this section, addressing the
research questions posed in the introduction.

@@

The Figure \ref{fig:graph1} shows that around 5\% of the prompts
written by students were not categorized. These messages range from
greetings to the Bot, such as a simple "\textit{Hola}," to
contextual statements like "\textit{Estoy repasando orientación a
objetos.}"

% Approximately 64\% of the messages pertain to categories associated with
% critical thinking, corroborating with the findings of
% \cite{Ghimire24}. In contrast, around 28\% of the messages
% indicate a preference among students for ready-made answers.

% Since most of the messages are located in categories where students are 
% notoriously more active in their learning (conceptual questions 32.1% and 
% Debugging Help 30.9%), it seems that students are trying to use the AI in a 
% way that promotes their learning. However, the pattern of conversations with 
% sequential blocks and also little spacing may suggest that there is a gap in 
% knowledge of study strategies.

\begin{figure*}[htbp]
  \centering
  \includegraphics[scale=0.55]{img/figure3.png}
  \caption{Interaction times per categories.}
  \label{fig:graph3}
\end{figure*}

@@

% Spacing
Spacing, a well-established learning strategy \citep{Carvalho20}, was examined
by analyzing the time intervals between users’ interactions with CharlieBot.
Although only 7.4\% of the conversations exhibited message delays exceeding 24
hours, 26.19\% involved response times longer than 60 minutes. These findings
suggest that students typically do not extend conversations across multiple days.
However, the data also reveal that some conversations had one or more pauses
exceeding an hour, a practice known to enhance knowledge retention.

% Another interesting study technique is spacing. Only 7.4% of the conversations
% exhibited message delays exceeding 24 hours and 26.19% involved response times 
% longer than 60 minutes. However, Carlos told me that this may be due to some 
% limitation of the bot, as it only maintains a conversation while the browser 
 % tab is open.

 % There is a big difference (I haven't looked at it statistically yet) in the 
%  interaction time with the Bot when a student asks conceptual questions. 
%  Perhaps this actually makes sense, since it is a category where, from the 
%  point of view of the metacognitive objective, the student is more concerned 
%  with understanding a concept and deeper learning.

@@

Intervealing is a study strategy that involves mixing or alternating different
topics or problems within a single learning session. This
approach has been shown to be useful for promoting long-term learning and
improving the application of knowledge in other situations \cite{Rivers21}.

When analyzing the classification of messages in conversations, we identified
the presence of sequences of messages belonging to the same category, that it
could suggest a low occurrence of interleaving. To better assess the extent of
these sequences, we calculated the percentage of messages that were part of
repeated blocks.

Considering blocks with at least three messages, we observed that 56.2\% of the
students' prompts were grouped into some sequence. When analyzing blocks with a
minimum of 4 messages, we found that 47.5\% of the prompts were part of
continuous sequences. These results indicate a moderate incidence of
interleaving among the students' messages.

To illustrate the presence of blocks the Figure \ref{fig:graph2} presents
examples of conversations and their classification. In Conversation 2, the
student alternates messages among Code Snippet, Debugging Help and Conceptual
Question, showing an tendency of interleaving between practical and theoretical
perspectives. In contrast, in Conversation 10, the student focused on a single
block of Conceptual Question. Although the focus was on understanding a
theoretical concept, this conversation tended not to promote interleaving, due
to the absence of alternation between different types of tasks.

% % Conceptual Questions
% \subsubsection*{Sequence of Conceptual Question}

% Messages classified as conceptual questions represent around 32\% of all
% messages. We have found a pattern of sequences of Conceptual Questions in some
% conversations. Some conversations are entirely composed of messages classified
% as Conceptual Questions. In other cases, the sequence of Conceptual questions
% starts with a request for a Complete Solution or a Code Snippet, followed by
% a progression of Conceptual Questions. In both cases, the student is actively
% engaging with the bot to understand the concepts and compromised with more
% deep learning.

@@

\section{Conclusion}

The results of this study suggest that students interact with CharlieBot in a
manner that promotes critical thinking. The majority of the messages analyzed
were related to debugging help, conceptual questions, and student correction,
indicating that students are actively engaging with the bot to understand
concepts and solve problems. This is a positive outcome, as it suggests that
CharlieBot is being used as a tool to enhance learning rather than a crutch for
students to rely on.

\section*{\uppercase{Appendix}}

>>>>

\title{
  Temporary: An Analysis of Programming Students’ Interaction Prompts with
  Generative AI: A Metacognitive Perspective
}

@@

\keywords{Generative Artificial Intelligence, Metacognition, Programming Education, Chatbots}

@@

Recent advancements in Generative Artificial Intelligence (GenAI) have opened up
new possibilities in education. In programming courses, students can utilize
GenAI tools to improve their understanding, receive feedback, and obtain
detailed explanations.

The study guided by \cite{chan23} revealed that both undergraduate and
postgraduate students exhibit positive attitudes toward the use of GenAI in
teaching and learning, noting that it enhances the depth of their thinking and
understanding. Additionally, a systematic review conducted by \cite{Lo24}
demonstrated that students could effectively learn from ChatGPT, leading to
improved understanding and academic achievement. Besides that, it was also
observed that ChatGPT enables students to regulate their learning pace
\citep{Baha24} and can potentially support students’ self-regulated learning,
especially for those with previous technical and disciplinary knowledge
\citep{Xia23}.

On the other hand, researchers have concerns about the impact of these tools on
students. The systematic review conducted by \cite{Murillo23} indicated that
ChatGPT use might lead to an overreliance on the tool. \cite{chan23} also noted
that this reliance could result in a decrease in critical thinking, as students
might make decisions based solely on the information provided by ChatGPT.

The student’s confidence is justified, as demonstrated by \cite{Puryear22}, who
found that GitHub Copilot can generate solutions for student assignments with
accuracy rates ranging from 68\% to 95\%. However, this also raises concerns
that students may become overly reliant on GenAI tools, potentially bypassing
the opportunity to develop a deeper understanding of underlying concepts.
\cite{cai23} further highlighted overdependence and diminished intellectual
engagement as significant drawbacks associated with the use of ChatGPT in
learning.

Regardless of teachers' preferences or beliefs, a preliminary surveys performed
by \cite{Dickey24} revealed that at least 54.5\% of students are already using
GenAI for homework. This highlights the need to increase the understanding of
how students are interacting with these tools and how they can be used to
enhance learning. Furthermore, a review organized by \cite{Lo24} emphasized the
necessity for extended studies and objective measures to gather more robust
evidence on the use of GenAI tools in education.

The primary objective of this study is to analyze the interactions between
students in a Java programming course and an educational bot powered by
ChatGPT 3.5, enhanced with Retrieval-Augmented Generation (RAG) capabilities,
within the context of two Metacognitive strategies: Interleaving and Spacing.
To achieve this objective, we formulated three research questions:

\begin{enumerate}
  \item What was the distribution pattern of the classified messages?
  \item Was there a space between student prompts? If so, which category promote the largest spacing between submissions?
  \item Do students' conversations with the bot alternate between practical and theoretical approaches, thereby promoting interleaving?
\end{enumerate}

@@

However, it is crucial to consider how GenAI tools impact students across
various demographic groups, academic disciplines, cultural backgrounds, and
levels of prior experience \citep{catalan21} \citep{neo22}. Consequently, this
study is limited to a specific group of students and focuses on the use of a
single GenAI tool. As a result, the findings may not be generalizable to other
populations or tools.

%% Are there any existing solutions?
% To deal with this issue, some researchers have proposed different pedagogical
% strategies \cite{Denny24} introduced the concept of \textit{Prompt
% Problems}, where students solve programming exercises by formulating natural
% language prompts. \cite{Prasad24} proposed a self-regulated learning framework
% using GenAI to solve programming problems. \cite{Lauren23} explored the
% integration of GenAI with evidence-based learning strategies in computer science
% and engineering courses.

%% Which is the best?

% The study of effective educational strategies in use of GenAI tools in education
% context are important for determining the best training for educators. However,
% as these tools gain popularity among students \cite{Dickey24} the urgency to
% equip educators with best practices may lag behind their rapid adoption.
% Therefore, analyzing students' interaction patterns with GenAI tools is
% important for understanding how these tools are being used and how educators can
% leverage them to enhance learning. In addition, these interaction patterns can
% create new opportunities to redesign GenAI tools to better support pedagogical
% strategies without repressing the development of abstract, critical and creative
% thinking.

@@

TODO: Work in progress

@@

% Regarding self-regulation, \cite{cai23} noted that ChatGPT allowed students to
% regulate their learning pace.

\begin{table*}[htbp]
  \caption{Categories - adapted from \cite{Ghimire24}}
  \begin{center}
    \renewcommand{\arraystretch}{1.6} % Increase the spacing between rows
    \begin{tabular}{p{3cm} p{12cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Description} \\
      \hline
      Debugging Help & Prompts that seek help to identify, fix errors, or understand the provided code snippet. \\
      Conceptual Question & Prompts that are more about understanding concepts than specific code. \\
      Student Correction & Prompts where the student corrects the bot. \\
      Code Snippet & Prompts that ask for a specific part of the code, like a function or a segment. \\
      Complete Solution & Prompts that request an entire solution or a complete code snippet. \\
      Multiple Question & Prompts where the user wants to solve a multiple choice exercise (Quiz). \\
      Language Change & Prompts that request a change of idiom. \\
      Uncategorized & Prompts that do not fit into any of the above categories. \\
      \hline
    \end{tabular}
    \label{tab:categories}
  \end{center}
\end{table*}

@@

Metacognitive knowledge includes beliefs about variables that affect the
outcomes of cognitive activities. It is divided into three types: beliefs about
personal abilities, perceived difficulty in the task, and previously used
strategies. Metacognitive Experience refers to the feelings that arise
before, during, and after cognitive activity, such as frustration, confusion,
and satisfaction, among others. Metacognitive Goals are the key to regulating
thought as they relate to the goals the individual seeks to achieve, directly
influencing the actions taken. For example, if a student’s goal is to complete
a task quickly, they may adopt a more passive approach to learning. Lastly,
Metacognitive Actions involve the planning, monitoring, and evaluation of
strategies used to achieve the goals. In terms of planning, students can
determine how to approach a task, such as spacing their study sessions
\citep{Ouhao18, Carvalho20}, interleaving topics \citep{Rivers21}, utilizing
retrieval practice \citep{larsen18}, and other strategies.

@@

In the categorization phase, the students' messages were manually classified
into eight distinct categories for analysis. Initially, \cite{Ghimire24}
proposed four categories: Debugging Help, Conceptual Question, Code Snippet,
and Complete Solution. However, the data collected indicated the need for
additional categories, leading to the inclusion of four more: Multiple Choice
(Quiz), Student Corrections, Language Change, and Uncategorized. In this study,
students were free to use the bot as they wished, which contributed to the
emergence of these additional categories. Table \ref{tab:categories} presents
these categories along with their respective descriptions.

The analysis phase involved the use of Python scripts to extract information
from previously classified data. The datasets and code snippets used in this
process are available at: YYY.

@@

A total of 343 students' messages were categorized across 42 conversations, with
an average of 8.17 messages per conversation. This data offers insights into how
users engage with CharlieBot and the types of prompts they submit. The analysis
results are presented in this section, addressing the research questions
outlined in the introduction.

@@

About the first research question: \textit{What was the distribution pattern of
the classified messages?.}

According to Figure \ref{fig:graph1} messages classified as
Conceptual Question account for 32.1\% of the total. Often, the sequence of
Conceptual Question begins with a request for a Complete Solution or Code
Snippet, which then evolves into a series of Conceptual Question, as seen in
conversation 2 of Figure \ref{fig:graph2}. In other cases, the conversation
consists entirely of messages classified as Conceptual Question, such as in
conversation 10 of Figure \ref{fig:graph2}. This type of message indicates that
students recognize they have not yet fully mastered a concept and are seeking to
better understand it. In this context, the student realizes that solving the
problem requires understanding the concepts, thereby applying an active learning
strategy.

As shown in Figure \ref{fig:graph1}, approximately 30.9\% of the messages were
classified as Debugging Help. These prompts often begin with a request for a
Complete Solution or Code Snippet. Following this, students typically shift
from a passive to a more active learning strategy, as illustrated by the start
of conversation 12 in Figure \ref{fig:graph2}. Debugging Help messages represent
a form of active learning, where students likely focus on gaining a practical
understanding of the code, identifying errors, and improving their debugging
skills for future tasks.

Since most of the messages are located in categories where students are
notoriously engaged in more active in their learning (Conceptual Question and
Debugging Help) corroborating with the findings of \cite{Ghimire24} which showed
that the prompts are also localized in the same categories

Requests for the bot to generate a Code Snippet or Complete Solution accounted
for 10.5\% of student messages. These prompts reflected a desire for more
direct answers; however, it was observed that, after receiving these solutions,
many students transitioned to a more active approach in seeking to understand
the code or the underlying concepts. Thus, it can be inferred that prompts
within the Code Snippet or Complete Solution category were often used as a
starting point for more in-depth study.

Multiple-choice exercise resolutions accounted for 16.9\% of the responses.
These prompts typically involved students submitting one or more exercises to
CharlieBot and requesting solutions. We identified sequences of multiple-choice
questions, as illustrated by conversations 8, 9, and 15 in \ref{fig:graph2}.
This less active learning approach can occur for various reasons, ranging from
the belief that reviewing solutions to similar questions enhances learning, to
issues such as negligence, poor planning, or other possible factors.

Student corrections to the bot’s responses accounted for 2.6\% of the
interactions. In most cases, these corrections occurred when the student already
had the correct answer to an exercise and identified an error in the bot’s
response, as illustrated in conversations 4 and 9 in Figure \ref{fig:graph2}.
These corrections highlight that, although the bot provides quick feedback, it
is not always accurate, and its errors can be difficult for students to
identify.

The Figure \ref{fig:graph1} also shows that around 5\% of the prompts written by
students were not categorized. These messages range from greetings to the Bot,
such as a simple "\textit{Hola}," to contextual statements like
"\textit{Estoy repasando orientación a objetos.}"

@@

% Spacing

About the second research question: \textit{Was there a space between student
prompts? If so, which category promotes the largest spacing between
submissions?}

Spacing, a well-established learning strategy \citep{Carvalho20}, was analyzed
by examining the time intervals between users' interactions with CharlieBot.
Only 9\% of the conversations showed message delays exceeding 24 hours,
indicating that students generally do not extend interactions across multiple
days. However, this students' behavior is directly influenced by the
implementation of the initial version of CharlieBot, which was designed to keep
the chat session active only while the student kept the browser tab open and
avoided reloading the page.

On the other hand, about 26.2\% of the conversations, or roughly a quarter,
included pauses longer than 60 minutes, a practice known to help with knowledge
retention. However, this data only shows that students paused their use of the
AI, which could mean they were reflecting on the topic independently or engaging
in unrelated activities, making it uncertain in what context the students were
applying spacing.

There is a noticeable difference in the interaction time with the bot when a
student asks a conceptual question. This was probably due to the fact that, in
this category, the student is more focused on understanding a concept and
engaging in deeper learning, which aligns with the metacognitive goal of
promoting comprehension and long-term retention. The Figure \ref{fig:graph3}
shows the average interaction time per category.

\begin{figure*}[htbp]
  \centering
  \includegraphics[scale=0.55]{img/figure3.png}
  \caption{Average interaction time per category.}
  \label{fig:graph3}
\end{figure*}

@@

Regarding the third research question: \textit{Do students' conversations with
the bot alternate between practical and theoretical approaches, thereby
promoting interleaving?}

Intervealing is a learning strategy that involves mixing or alternating
different topics or problems within a single learning session. This approach has
been shown to be useful for promoting long-term learning and improving the
application of knowledge in other situations \citep{Rivers21}.

When analyzing the classification of messages in conversations, we identified
the presence of sequences of messages belonging to the same category, that it
could suggest a low occurrence of interleaving. To illustrate the presence of
blocks the Figure \ref{fig:graph2} presents examples of conversations and their
classification. In Conversation 2, the student alternates messages among Code
Snippet, Debugging Help and Conceptual Question, showing an tendency of
interleaving from practical to theoretical perspective. In contrast, in
Conversation 10, the student focused on a single block of Conceptual Question.
Although the focus was on understanding a theoretical concept, this conversation
tended not to promote interleaving, due to the absence of alternation between
different types of tasks.

To better assess the extent of these sequences, we calculated the percentage of
messages that were part of repeated blocks. Considering blocks with at least
three messages, we observed that 56.2\% of the students' prompts were grouped
into some sequence. When analyzing blocks with a minimum of four messages, we
found that 47.5\% of the prompts were part of continuous sequences.

However, if we consider that sequential messages from a certain point indicate
that the student is \textit{blocked} (or focused) on a task, the remaining space
for interleaving is 43.8\%, assuming a block starting from at least three
consecutive messages, and 52.5\% for blocks starting from four consecutive
messages. Thus, the data suggest that interleaving occurs on students'
conversations with CharlieBot but in which situations and with what type of
students profile is a study that needs to be further investigated.

\section{Conclusion and Future Work}

TODO: Work in progress

* Although the data indicates some use of interleaving and spacing strategies,
we believe that students can be guided to adopt a more deliberate approach in
utilizing GenAI, enabling them to more effectively harness these tools to
enhance their learning.

* Due to the anonymity of the interactions, it's unclear whether
students were intentionally using interleaving and spacing strategies or if
these patterns emerged naturally. Future research should investigate whether
students were consciously applying.

* To provide teachers with valuable data for analyzing how students are using
the tools, we are developing an AI model to automatically classify student
messages. This will enable educators to better understand student behavior and
offer more targeted support where needed.

\section*{\uppercase{\textit{Appendix}}}

% % Conceptual Question
% \subsubsection*{Sequence of Conceptual Question}

% Messages classified as conceptual question represent around 32\% of all
% messages. We have found a pattern of sequences of Conceptual Question in some
% conversations. Some conversations are entirely composed of messages classified
% as Conceptual Question. In other cases, the sequence of Conceptual question
% starts with a request for a Complete Solution or a Code Snippet, followed by
% a progression of Conceptual Question. In both cases, the student is actively
% engaging with the bot to understand the concepts and compromised with more
% deep learning.

@@
